Confirmatory factor analysis of the Maltreatment and Abuse Chronology of Exposure (MACE) scale: Evidence for essential unidimensionality

### 2022-09-20 manuscript/main.tex
ready for submission

\documentclass[letterpaper,man,natbib,noextraspace,floatsintext,longtable]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}             % fancy latex tables
\usepackage[export]{adjustbox}    % center wide tables on page
\usepackage{setspace}             % adjust caption linespacing 
\usepackage{multibib}             % references in supplement
\usepackage{hyperref}             % manage hyperlink colors

@@

\title{Confirmatory factor analysis of the Maltreatment and Abuse Chronology of Exposure (MACE) scale: Evidence for essential unidimensionality}
\shorttitle{Confirmatory factor analysis of the MACE}
\author{Samuel Zorowitz$^1$, Lauri Tuominen$^{2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$The Royal’s Institute of Mental Health Research, University of Ottawa, Canada}

\abstract{

\noindent \textbf{Background}: The Maltreatment and Abuse Chronology of Exposure (MACE) scale is a retrospective self-report scale that measures 10 distinct types of childhood maltreatment. Despite its increasing use, the factor structure of the MACE has not been thoroughly investigated. As such, the viability of the MACE total and subscale scores are uncertain. The aim of the present study was to investigate the factor structure of the MACE and to quantify the reliability of its total and subscale scores.

@@

\noindent \textbf{Methods}: We evaluated a series of confirmatory item factor models fit to the responses of two independent samples of participants (N=1051 \& N=582) who completed the MACE. The models included hierarchical (bifactor) models. We used model-based indices to estimate the reliability of the MACE total and subscale scores, and to quantify the essential unidimensionality of the scale.

@@

\noindent \textbf{Results}: We found the MACE total score was a reliable and valid measure of overall childhood maltreatment. In contrast, although MACE subscale scores exhibited adequate reliability, we found that the vast majority of reliable variance in these scores reflected general maltreatment and not any particular type of maltreatment as intended. We also found that the MACE is essentially unidimensional, and exhibits evidence of differential item functioning by gender.

@@

\noindent \textbf{Conclusions}: Our results provide support for a one-factor structure of the MACE, as well as continued use of the MACE total score. Our results caution against the use of MACE subscale scores, which have little practical use given that they provide little unique or reliable information above and beyond the total score.

\hfill \break

@@

\authornote{This research was funded by the Royal's Emerging Research Innovators in Mental Health program at the University of Ottawa Institute of Mental Health Research and a National Science Foundation Graduate Research Fellowship. The authors wish to thank Dr. Robyn McQuaid for thoughtful feedback on this manuscript. The authors have no conflicts of interest to declare.}

@@

There already exist many measures of childhood abuse and maltreatment for researchers to choose from \citep{saini2019systematic}. One recently developed measure, the Maltreatment and Abuse Chronology of Exposure scale (MACE; \citealt{teicher2015maltreatment}), has a number of notable advantages. First, the MACE is a 52-item retrospective self-report scale composed of 10 subscales, each of which measures a unique type of childhood maltreatment. The content of six of these subscales overlap with other childhood maltreatment scales (e.g. physical abuse, neglect), whereas the remaining four are less commonly represented in other scales (e.g. peer victimization, witnessing domestic violence). Second, the items in the MACE were carefully selected to measure maltreatment of increasing severity in order to finely differentiate people with differing exposure levels. Third, the MACE also asks for the onset and duration of maltreatment experiences, which is necessary for investigating how the timing of abuse shapes development. Finally, the MACE was originally developed to provide two measures: a total score (number of overall maltreatment experiences) and a multiplicity score (number of types of maltreatment experienced). These scores have excellent temporal stability, convergent validity with other childhood maltreatment scales, and good predictive validity for a multitude of psychiatric symptoms \citep{teicher2015maltreatment}. For these reasons, the MACE has been included as among the best measures of childhood maltreatment currently available \citep{saini2019systematic, georgieva2022systematic}.

In recent years, there have been calls for childhood maltreatment research to move away from the cumulative risk framework \citep{evans2013cumulative} --- which focuses only on the number of adverse experiences --- towards a dimensional approach, which emphasizes how distinct types of maltreatment uniquely shape developmental trajectories \citep{mclaughlin2016beyond, belsky2012beyond}. In response, some researchers have begun to use the MACE to make new indices of maltreatment in lieu of, or in addition to, the total and multiplicity scores. For example, some researchers have used all 10 MACE subscale scores to study the unique contributions of each type of maltreatment to the development of psychopathology \citep{schalinski2015type, gerke2018childhood, schalinski2019early}. Citing a prominent dimensional model of maltreatment \citep{mclaughlin2014childhood}, others have made novel threat and deprivation subscale scores made up of items in the MACE that measure abuse and neglect experiences, respectively \citep{schalinski2018defining, schalinski2019environmental, teicher2018differential}. The psychometric properties of these subscale scores have not been thoroughly investigated.

@@

To investigate the factor structure of the MACE, we fit a series of four confirmatory item factor models separately to the response data from each sample.\footnote{The suitability of common factor models (CFMs) with reflective indicators for explaining childhood maltreatment data has been contested. For interested readers, we justify our use of CFMs to explain responses to the MACE in the supplement.} The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each model is depicted in Figure \ref{fig:models}. The first is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b). This model will be used to evaluate if the MACE can produce reliable subscale scores for each of the 10 types of maltreatment assessed by the MACE.

@@

We also identified a reverse-scored methods factor in the MACE. In our view, the most plausible explanation is that, under the \cite{teicher2015maltreatment} scoring procedure, affirmative responses on the regular and reverse-scored items on the MACE represent very different endorsements. For a regular item on the MACE, an endorsement indicates that a person experienced an adverse event during one year of childhood at least. In contrast, an endorsement on a reverse-scored item indicates that a person experienced experienced the absence of a positive event for all 18 years of childhood. However, alternative causes of the reverse-scored factor (e.g. participant confusion, acquiescence, or carelessness; \citealt{weijters2013reversed}) are possible. Additional research is needed to see if alternative scoring procedures would eliminate this factor.

>>>>

\documentclass[letterpaper,man,natbib,noextraspace,floatsintext,longtable,12pt]{apa6}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}             % fancy latex tables
\usepackage[export]{adjustbox}    % center wide tables on page
\usepackage{setspace}             % adjust caption linespacing 
\usepackage{multibib}             % references in supplement
\usepackage{hyperref}             % manage hyperlink colors
\usepackage{hanging}

@@

% define title page
\title{Confirmatory factor analysis of the Maltreatment and Abuse Chronology of Exposure (MACE) scale: Evidence for essential unidimensionality}
\shorttitle{Confirmatory factor analysis of the MACE}
\author{Samuel Zorowitz$^{1*}$, Lauri Tuominen$^{2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$The Royal’s Institute of Mental Health Research, University of Ottawa, Canada}

% add author note
\authornote{

\noindent *Corresponding author: zorowitz@princeton.edu

{\begin{hangparas}{.25in}{1} Acknowledgments: The authors wish to thank Dr. Robyn McQuaid for thoughtful feedback on this manuscript. \end{hangparas}}

{\begin{hangparas}{.25in}{1} \noindent Ethics approval: The Institutional Review Board of the University of Ottawa (\#2021003) approved this study. \end{hangparas}}

{\begin{hangparas}{.25in}{1} \noindent Funding sources: This research was funded by the Royal's Emerging Research Innovators in Mental Health program at the University of Ottawa Institute of Mental Health Research and a National Science Foundation Graduate Research Fellowship. \end{hangparas}}

\noindent Conflict of interest: The authors have no conflicts of interest to declare.

% add abstract
\abstract{

\noindent \textbf{Background}: The Maltreatment and Abuse Chronology of Exposure (MACE) scale is a retrospective self-report scale that measures 10 distinct types of childhood maltreatment. Despite its increasing use, the factor structure of the MACE has not been thoroughly investigated. As such, the viability of the MACE total and subscale scores are uncertain.

@@

\noindent \textbf{Objectives}: The current study investigated the factor structure of the MACE in order to quantify the reliability of its total and subscale scores.

@@

\noindent \textbf{Participants and setting}: Two independent samples of participants (N=1051 \& N=582) who completed the MACE were included in this study.

@@

\noindent \textbf{Methods}: Using confirmatory item response models, we tested one-factor and several bifactor models of participants' responses. We used model-based indices to estimate the reliability of the MACE total and subscale scores, and to quantify the essential unidimensionality of the scale.

\noindent \textbf{Results}: We found the MACE total score was a reliable and valid measure of overall childhood maltreatment. In contrast, although we found MACE subscale scores exhibited adequate reliability, the vast majority of their reliable variance reflected general maltreatment and not any particular type of maltreatment as intended. We also found that the MACE is essentially unidimensional and some of its items exhibit of differential item functioning by gender.

\noindent \textbf{Conclusions}: Our results provide support for a one-factor structure of the MACE, as well as continued use of the MACE total score. Our results caution against the use of MACE subscale scores, which have little practical use insofar that they provide little unique, reliable information above and beyond the total score.

{\begin{spacing}{1.15} \hfill \\ \end{spacing}}

@@

@@

There already exist many measures of childhood abuse and maltreatment for researchers to choose from \citep{saini2019systematic}. One recently developed measure, the Maltreatment and Abuse Chronology of Exposure scale (MACE; \citealt{teicher2015maltreatment}), has a number of notable advantages. First, the MACE is a 52-item retrospective self-report scale composed of 10 subscales, each of which measures a unique type of childhood maltreatment. The content of six of these subscales overlap with other childhood maltreatment scales (e.g. physical abuse, neglect), whereas the remaining four are less commonly represented in other scales (e.g. peer victimization, witnessing domestic violence). Second, using item response theory, the items in the MACE were carefully selected to measure maltreatment experiences of increasing severity in order to finely differentiate people with differing exposure levels. Third, the MACE also asks for the onset and duration of maltreatment experiences, which is necessary for investigating how the timing of abuse shapes development. Finally, the MACE was originally developed to provide two measures: a total score (number of overall maltreatment experiences) and a multiplicity score (number of types of maltreatment experienced). These scores have excellent temporal stability, convergent validity with other childhood maltreatment scales, and good predictive validity for a multitude of psychiatric symptoms \citep{teicher2015maltreatment}. For these reasons, the MACE has been included as among the best measures of childhood maltreatment currently available \citep{saini2019systematic, georgieva2022systematic}.

In recent years, there have been calls for childhood maltreatment research to move away from the cumulative risk framework \citep{evans2013cumulative} --- which focuses only on the number of adverse experiences --- towards a dimensional approach, which emphasizes how distinct types of maltreatment uniquely shape developmental trajectories \citep{mclaughlin2016beyond, belsky2012beyond}. In response, some researchers have begun to use the MACE to make new indices of maltreatment in lieu of, or in addition to, the total and multiplicity scores. For example, some researchers have used all 10 MACE subscale scores to study the unique contributions of each type of maltreatment to the development of psychopathology \citep{schalinski2015type, gerke2018childhood, schalinski2019early}. Citing a prominent dimensional model of maltreatment \citep{mclaughlin2014childhood}, others have made novel threat and deprivation subscale scores made up of items in the MACE that measure abuse and neglect experiences, respectively \citep{schalinski2018defining, schalinski2019environmental, teicher2018differential}. However, the psychometric properties of these subscale scores have not been thoroughly investigated.

@@

To investigate the factor structure of the MACE, we fit a series of four confirmatory item factor models separately to the response data from each sample. (The suitability of common factor models (CFMs) with reflective indicators for explaining childhood maltreatment data has been contested. For interested readers, we justify our use of CFMs to explain responses to the MACE in the supplement.) The basis of each model is the graded response model for ordered polytomous data \citep{samejima1997graded}. The factor structure for each model is depicted in Figure \ref{fig:models}. The first is a one-factor (undimensional) model where all items load onto a general maltreatment factor (Figure \ref{fig:models}a). The second is a bifactor model where all items load onto a general maltreatment factor and also one of 10 specific factors based on their subscale membership (Figure \ref{fig:models}b). This model will be used to evaluate if the MACE can produce reliable subscale scores for each of the 10 types of maltreatment assessed by the MACE.

@@

We identified 12 item pairs and triplets in the MACE with response dependence; that is, sets of items where a response to one item requires or strongly implies a certain response to another. These items violate the local independence assumption of item response theory models and, in the present analyses, were therefore collapsed into polytomous items. However, future studies might consider abbreviating the MACE by removing some of these dependent or redundant items altogether (possibly without a significant loss of test information). We also identified a reverse-scored methods factor in the MACE. In our view, the most plausible explanation is that, under the \cite{teicher2015maltreatment} scoring procedure, affirmative responses on the regular and reverse-scored items on the MACE represent very different endorsements. For a regular item on the MACE, an endorsement indicates that a person experienced an adverse event during one year of childhood at least. In contrast, an endorsement on a reverse-scored item indicates that a person experienced the absence of a positive event for all 18 years of childhood. However, alternative causes of the reverse-scored factor (e.g. participant confusion, acquiescence, or carelessness; \citealt{weijters2013reversed}) are possible. Additional research is needed to see if alternative scoring procedures would eliminate this factor.
